---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=FALSE}
act<- read.csv("activity.csv")
print(colnames(act))
act$date <- as.Date(act[,"date"], format="%Y-%m-%d")
act <- na.omit(act)
days <- split(act, act$date)

```

## What is mean total number of steps taken per day?
```{r steps per day calculations}
steps_per_day <- unlist(lapply(days, function(x){ sum(x$steps)}))
mean_steps_per_day <- as.integer(mean(steps_per_day))
median_steps_per_day <- median(steps_per_day)
```

The mean and the median of the total number of steps taking per days are `r mean_steps_per_day` and `r median_steps_per_day` respectivelly.

```{r, echo=TRUE}
hist(steps_per_day, breaks = length(days)/2, xlab= "Number of steps per day", col = "blue", main = "Histogram of the steps per day")
abline(v = mean_steps_per_day,  col="red")
abline(v = median_steps_per_day, col="green")
```


## What is the average daily activity pattern?
Make a time series plot (i.e. type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
min_5_interval <- split(act, act$interval)
aver_min_5_interval <- unlist(lapply(min_5_interval, function(x){ mean(x$steps)}))

```

```{r, echo=TRUE}
plot(aver_min_5_interval, xlab = "5 minutes interval", ylab="Average number of steps", type= "l",  col = "blue", main = "Averaged of steps taking every 5 min across all days ")

```

## Imputing missing values
Calculate de total number of rows with NA

```{r, echo=TRUE}
act_tot<- read.csv("activity.csv")
#numb NAs are dim_tot minus dim activities where na.omit was used
numb_NAs = dim(act_tot)[1]-dim(act)[1]
```

The total number of rows with NAs are: `r numb_NAs`.

Filling the NAs steps with the mean of the 5 min interval and create a new database that have them replace.

```{r "filling NAs with mean per 5 min", echo=TRUE}
act_tot$steps <- as.numeric(unlist(apply(act_tot, MARGIN = 1, function(x){
    if(is.na(x[1])){
        int<-gsub("\\s", "", x[3])
        x[1] <- aver_min_5_interval[names(aver_min_5_interval)==int]
    }
return(x[1])
})))
```

```{r "steps per day calculations w NAs filled", echo=TRUE}

act_tot$date <- as.Date(act_tot$date)#, "%Y-%m-%d")

days_n <- split(act_tot, act_tot$date)

steps_per_day_n <- unlist(lapply(days_n, function(x){ sum(x$steps)}))

mean_steps_per_day_n <- as.integer(mean(steps_per_day_n))
median_steps_per_day_n <- as.integer(median(steps_per_day_n))
```

The mean and the median of the total number of steps per days in the database with the NAs replaced by the average of the correspondind 5 min intervals are `r mean_steps_per_day_n` and `r median_steps_per_day_n` respectivelly.

```{r, echo=TRUE}
hist(steps_per_day_n, breaks = length(days_n)/2, xlab= "Number of steps per day", col = "blue", main = "Histogram of the steps per day with NAs replaced by mean of 5 min interval")
abline(v = mean_steps_per_day_n,  col="red")
abline(v = median_steps_per_day_n, col="green")
```

By looking at the both histograms with and without NAs there is not a much difference with respect to the shape and also the mean and the median do not change to much.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays, echo=TRUE}
Sys.setlocale("LC_TIME", "C")

act_tot$date <- as.Date(act_tot$date)#, "%Y-%m-%d")
weekdays_n<- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
act_tot$week_days <- factor((weekdays(act_tot$date) %in% weekdays_n), 
         levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
```

```{r, echo=TRUE}
steps_per_day_n <- aggregate(steps~interval + week_days, data=act_tot, mean)
print(steps_per_day_n[1:3,])

```

```{r, echo=TRUE}
library(lattice)
xyplot(steps ~ interval | week_days, data=steps_per_day_n, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")
```

